---
title: "4_pregnancy"
author: "Gabrielle Goldet"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
libraries and functions

```{r}
library(dplyr)
library(lubridate)

#filtering according to whether an individual has or doesn't have more than 1 patid
single <- function(df){df %>% filter(is.na(individual_id))}
multiple <- function(df){df %>% filter(!is.na(individual_id))}


#propagating relevant data across all the patids of a given individual
fill_columns <- function(df, ...) {
  df %>% 
    left_join(all_pats, by = "individual_id") %>%
    group_by(individual_id) %>%
    fill(..., .direction = "downup") %>%
    ungroup()
}
```

This file walks through how to determine delivery dates from the pregnancy observations.  the pregnancy code is flagged for observations that indicate a delivery.  Note also that a delivery date can be imputed by backdating the observation date of the maternal 6 weeks postnatal examination by 6 weeks.

Start by gatethering all the pregnancy observations then all the delivery observations. If you have all your obersvation files in a list of dfs (observation_list - in our case we compounded all the 200+ observation files into 17 dfs which are in this list), you can select all rows related to pregnancy observations 
```{r}
setwd("YOUR DIRECTORY")
pregnancy_code <- read.delim("pregnancy_code.txt", stringsAsFactors = TRUE)
pregnancy_ICD10_code <- read.delim("pregnancy_ICD10_code.txt", stringsAsFactors = TRUE)
all_HES_apc <- #load your hes apc data
#also load your data - in our case it is stored in observation_list and all the HES APC data

pregnancy_fun <- function(df){df %>% inner_join(pregnancy_code, by = "medcodeid") %>% filter(is.na(obsdate))} # also ensure to remove observations without a date

#get all cprd codes
all_pregnancy_obs <- bind_rows(lapply(observation_list, pregnancy_fun))

#get all your HES data
all_hes_pregnancy <- all_HES_apc %>% 
  inner_join(pregnancy_ICD10_code, by = "ICD")# note delivery codes are flagged in this code list too

# now load all hes_maternity data if you have it
hes_maternity$epistart <- as.Date(hes_maternity$epistart, format = "%Y/%m/%d")

hes_deliveries <- hes_maternity %>% 
    filter(numbaby=="1" | numbaby=="2"| numbaby=="3"| numbaby=="4") %>%
  distinct(patid, epistart) %>% 
  mutate(delivery = 1) %>% 
  rename(obsate=epistart)


#now bind all pregnancy data
all_pregnancy_obs <- bind_rows(all_pregnancy_obs, hes_maternity, hes_deliveries)
```

first we need to add the individual ids of those with mltiple patids and also the years of birth so that dates of pregnancy that are likely dated erroneously (i.e before the age of, say, 14) are removed
.  Note there is a delivery colum in delivery obs that comes from the pregnancy_code.  This flags delivery observations
```{r}
all_pregnancy_obs <- all_pregnancy_obs %>% 
  left_join(patient_file %>% select(patid, yob), by = "patid") %>% 
  left_join(all_pats, by = "patid")

#filter the delivery obs and backdate observations related to a postnatal 6 week exam
delivery_obs <- all_pregnancy_obs_update %>% 
  filter(!is.na(delivery)) %>% 
  mutate(update_del_date = if_else(grepl("postnatal 6 week", Term, ignore.case = TRUE), obsdate %m-% weeks(6), NA)) %>% 
  mutate(obsdate = coalesce(update_del_date, obsdate)) %>% 
  filter(year(obsdate) > yob + 14)
```
now determine unique deliveries - look for clusters of delivery observations within a 2 month period separated by at least 9 months from the next delivery observations.  Do this separately for those with single and multiple patids

We identified individual delivery events using a two-stage approach to account for both data quality and biological plausibility. 
First, observation records were grouped into clusters within each patient, where any records occurring within 60 days of each other were assumed to relate to the same delivery episode; only the earliest date within each cluster was retained to reduce duplication from multiple records of a single event. 
Second, these cleaned episode-level dates were further evaluated using a minimum gap of 273 days (approximately nine months) to define distinct deliveries, ensuring that separate delivery events were biologically plausible. The earliest date within each resulting delivery group was then taken as the final delivery date. This approach allows us to minimise noise from repeated records while accurately identifying unique delivery events.
```{r}
#single patid
unique_deliveries_single <- single(delivery_obs) %>%
  arrange(patid, obsdate) %>%
  group_by(patid) %>%
  mutate(
    obsdate = as.Date(obsdate),
    gap_from_prev = as.numeric(obsdate - lag(obsdate)),
    # Create cluster IDs for episodes within 2 months (<= 60 days)
    new_cluster = if_else(is.na(gap_from_prev) | gap_from_prev > 60, 1L, 0L),
    cluster_id = cumsum(new_cluster)
  ) %>%
  ungroup() %>% 
  # Collapse each cluster to the earliest observation date
  group_by(patid, cluster_id) %>%
  slice_min(order_by = obsdate, n = 1, with_ties = FALSE) %>% #pick earliest date within the flucser
  ungroup() %>%
  arrange(patid, obsdate) %>%
  group_by(patid) %>%
  mutate(
    # Apply the 9-month rule between clustered delivery dates
    gap_from_prev_delivery = as.numeric(obsdate - lag(obsdate)),
    new_delivery = if_else(is.na(gap_from_prev_delivery) | gap_from_prev_delivery >= 273, 1L, 0L),
    delivery_id = cumsum(new_delivery)
  ) %>%
  ungroup() %>% 
  group_by(patid, delivery_id) %>% 
  slice_min(order_by = obsdate, n = 1, with_ties = FALSE) %>% #then pick the earliest date within each delivery_id
  ungroup()



unique_deliveries_multiple <- multiple(delivery_obs) %>%
  arrange(individual_id, obsdate) %>%
  group_by(individual_id) %>%
  mutate(
    obsdate = as.Date(obsdate),
    gap_from_prev = as.numeric(obsdate - lag(obsdate)),
    # Create cluster IDs for episodes within 2 months (<= 60 days)
    new_cluster = if_else(is.na(gap_from_prev) | gap_from_prev > 60, 1L, 0L),
    cluster_id = cumsum(new_cluster)
  ) %>%
  ungroup() %>% 
  # Collapse each cluster to the earliest observation date
  group_by(individual_id, cluster_id) %>%
  slice_min(order_by = obsdate, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(individual_id, obsdate) %>%
  group_by(individual_id) %>%
  mutate(
    # Apply the 9-month rule between clustered delivery dates
    gap_from_prev_delivery = as.numeric(obsdate - lag(obsdate)),
    new_delivery = if_else(is.na(gap_from_prev_delivery) | gap_from_prev_delivery >= 273, 1L, 0L),
    delivery_id = cumsum(new_delivery)
  ) %>%
  ungroup()%>% 
  group_by(individual_id, delivery_id) %>% 
  slice_min(order_by = obsdate, n = 1, with_ties = FALSE) %>% #then pick the earliest date within each delivery_id
  ungroup() %>% 
  select(-patid) %>% 
  distinct() %>% 
  fill_columns(obsdate, medcodeid, Term, ICD, delivery_id) ##expect many to many because any indvidiual could have more than one delivery

#bind all the deliveries
unique_deliveries <- bind_rows(unique_deliveries_multiple, unique_deliveries_single)

```
Now identify deliveries related to a GDM pregnancy.   First, note some codes in the pregnancy observation suggest PRE-EXISTING DM - any GDM during or after such a pregnancy must be ignored as this is likely DM
```{r}

all_pregnancy_obs <- all_pregnancy_obs %>% 
  mutate(preexisting = if_else(grepl("pre-existing type", Term, ignore.case = TRUE)|grepl("pre-existing diab", Term, ignore.case = TRUE), 1, NA))

#for single patid
earliest_preexisting_single <- single(all_pregnancy_obs) %>% 
  filter(!is.na(preexisting)) %>% 
  group_by(patid) %>% 
  slice_min(obsdate) %>% 
  slice(1) %>% 
  ungroup() %>% 
  rename(earliest_preexisting= obsdate)

#for multiple patid
earliest_preexisting_multiple <- multiple(all_pregnancy_obs_update) %>% 
  select(-earliest_preexisting) %>% 
  filter(!is.na(preexisting)) %>% 
  group_by(individual_id) %>% 
  slice_min(obsdate) %>% 
  slice(1) %>% 
  ungroup() %>% 
  rename(earliest_preexisting= obsdate) %>%
  select(-patid) %>% 
  fill_columns() #propagate across all the patids for an individual

#bind them together to facilitate left join
earliest_preexisting <- bind_rows(earliest_preexisting_single, earliest_preexisting_multiple) %>% 
  distinct(patid, earliest_preexisting)

all_pregnancy_obs_update <- all_pregnancy_obs%>% 
  left_join(earliest_preexisting %>% select(patid, earliest_preexisting), by = "patid") 

```

Now find valid gdm observations before any evidence of pre-existing DM and search for related delivery date.  Note that in the final step of assigning a patient to a particular group (GDM or non-GDM), only GDM deliveries predating the DM delivery date by at least a year will be counted as evidence of GDM so the process below is only one step in assuring that the GDM assignation is appropriate
```{r}
#check for valid GDM obs, ie before those suggesting chronic DM by at least a year
valid_gdm_obs_single <- single(all_pregnancy_obs_update) %>% 
  filter(obsdate < earliest_preexisting %m-% years(1)|is.na(earliest_preexisting)) %>% 
  filter(medcodeid %in% gdm_code$medcodeid|ICD=="O24.4") %>% #filter by those on the gdm medcodeid list or with the relevant ICD 10 code
  filter(!grepl("history", Term, ignore.case = TRUE)|is.na(Term)) %>% #remove the historic codes because we don't have a reliable observation date for them
  filter(!grepl("h/o", Term, ignore.case = TRUE)) #remove the historic codes because we don't have a reliable observation date for them

#now look for delivery dates related to those gdm episodes
gdm_pregnancy_search_single <- single(unique_deliveries) %>% 
  distinct(patid, obsdate, delivery_id) %>% 
  rename(delivery_date = obsdate) %>% 
  left_join(valid_gdm_obs_single %>% select(patid, obsdate), by = "patid") %>%   ##expect many-to-many warning
  filter(obsdate<= delivery_date %m+% months(6) & obsdate >= delivery_date %m-% months(9)) %>%  #look for a gdm observation within a reasonable time span ie within 6 months post partum or within 9 months of delivery
  distinct() %>% 
  group_by(patid, delivery_id) %>% 
  slice_min(obsdate) %>% 
  ungroup()

#now repeat the process for those with mutliple patids
valid_gdm_obs_multiple <- multiple(all_pregnancy_obs_update) %>% 
  filter(obsdate < earliest_preexisting %m-% years(1)|is.na(earliest_preexisting)) %>%
  filter(medcodeid %in% gdm_code$medcodeid|ICD=="O24.4") %>% 
  filter(!grepl("history", Term, ignore.case = TRUE)) %>% 
  filter(!grepl("h/o", Term, ignore.case = TRUE))

gdm_pregnancy_search_multiple <-multiple(unique_deliveries) %>% 
  distinct(individual_id, obsdate, delivery_id) %>% 
  rename(delivery_date = obsdate) %>% 
    left_join(valid_gdm_obs_multiple %>% select(individual_id, obsdate), by = "individual_id") %>%   ##expect many-to-many
  filter(obsdate<= delivery_date %m+% months(9) & obsdate >= delivery_date %m-% months(9)) %>% 
  distinct() %>% 
  group_by(individual_id, delivery_id) %>% 
  slice_min(obsdate) %>% 
  ungroup()
```
Now we have to go hunt for GDM codings outside of pregnancy episodes defined by the delivery dates created above.  For the next few code chunks we are just going to look at those with single patids.

```{r}
#rename obsdate to avoid confusion
unique_deliveries_single <- single(unique_deliveries) %>% 
  rename(delivery_date = obsdate)

#define start and end of pregnancy
unique_deliveries_single <- unique_deliveries_single %>%
        mutate(
          start = delivery_date %m-% months(9),  # 9 months before delivery
          end   = delivery_date %m+% months(3)   # 3 months after delivery
        )

  #find all obs during pregnancy
  in_pregnancy_single <- all_pregnancy_obs_update %>%
    inner_join(unique_deliveries_single %>% select(patid, start, end),#expect many to many
               by = "patid") %>%
    filter(obsdate >= start & obsdate <= end)
  
 #label the GDM obs in all_pregnancy_obs_update based on their presence on the gdm cprd code list or the relevant ICD 10 code
  all_pregnancy_obs_update <- all_pregnancy_obs_update %>% 
    mutate(gdm = if_else(medcodeid %in% gdm_code$medcodeid| ICD == "O24.4", 1, NA))
  
   #find gdm codings outside of pregnancy
 gdm_obs_outside_del_single <- single(all_pregnancy_obs_update) %>%
    anti_join(in_pregnancy_single %>% select(patid, obsdate, start, end), by = c("patid", "obsdate")) %>% #add start and end dates of already identified pregnancies
    group_by(patid) %>%
    filter(any(gdm == 1)) %>% #keep any patids that have a gdm coding
    ungroup() %>% 
        filter(obsdate < earliest_preexisting %m-% years(1)|is.na(earliest_preexisting)) %>% #remove pre-existing diabetes episodes
    filter(!grepl("history", Term, ignore.case = TRUE)) %>% #remove historical gdm codings
    filter(!grepl("h/o", Term, ignore.case = TRUE))
  
```
So now we have GDM codings outside of pregnancy.  Examining these it is clear that there are some gdm episodes that do actually have a delivery date but, for example, it was misdated as 01-01 or there is a historical delivery observation during the pregnancy that is conflicting.
Having removed the delivery date obtained above, if we cycle through to find more delivery dates that should remove some ambiguity, so we effectively need to repeat the search for delivery dates with the data outside the dates that were already found

```{r}
#now find new delivery dates for those with gdm episodes outside those previously found
 unique_deliveries_gdm_single <- gdm_obs_outside_del_single %>%
   filter(!is.na(delivery)) %>% 
   arrange(patid, obsdate) %>%
   group_by(patid) %>%
   mutate(
     obsdate = as.Date(obsdate),
     gap_from_prev = as.numeric(obsdate - lag(obsdate)),
     # Create cluster IDs for episodes within 2 months (<= 60 days)
     new_cluster = if_else(is.na(gap_from_prev) | gap_from_prev > 60, 1L, 0L),
     cluster_id = cumsum(new_cluster)
   ) %>%
   ungroup() %>% 
   # Collapse each cluster to the earliest observation date
   group_by(patid, cluster_id) %>%
   slice_min(order_by = obsdate, n = 1, with_ties = FALSE) %>%
   ungroup() %>%
   arrange(patid, obsdate) %>%
   group_by(patid) %>%
   mutate(
     # Apply the 9-month rule between clustered delivery dates
     gap_from_prev_delivery = as.numeric(obsdate - lag(obsdate)),
     new_delivery = if_else(is.na(gap_from_prev_delivery) | gap_from_prev_delivery >= 273, 1L, 0L),
     delivery_id = cumsum(new_delivery)
   ) %>%
   ungroup() %>% 
   group_by(patid, delivery_id) %>% 
   slice_min(order_by = obsdate, n = 1, with_ties = FALSE) %>% #then pick the earliest date within each delivery_id
   ungroup()


#define pregnancies by start and end date
 unique_deliveries_gdm_single <- unique_deliveries_gdm_single %>%
   rename(delivery_date = obsdate) %>% 
   mutate(
     start = delivery_date %m-% months(9),  # 9 months before delivery
     end   = delivery_date %m+% months(3)   # 3 months after delivery #give a bit of leeway for late coding)
   )
```
Now label the gdm pregnancies found in this round
```{r}

 #find further pregnancies associated with a delivery as defined by the second round
 in_pregnancy_gdm_single <- gdm_obs_outside_del_single %>%
   inner_join(unique_deliveries_gdm_single %>% select(patid, start, end),#expect many to many
              by = "patid") %>%
   filter(obsdate >= start & obsdate <= end) %>% 
      rename(delivery_date = obsdate) %>% 
   distinct(patid, delivery_date, start)
```
And now find all the gdm observations where we really don't have a delivery date, and bind all the relevant dfs to find all the gdm episodes
```{r}
 #now find all gdm data outside those deliveries
 gdm_obs_outside_del_single_round2 <- gdm_obs_outside_del_single %>%
   rename(delivery_date = obsdate) %>% 
   anti_join(in_pregnancy_gdm_single %>% select(patid, delivery_date, start), by = c("patid", "delivery_date")) %>% 
   filter(!is.na(gdm)) %>% 
   group_by(patid) %>% 
   slice_min(delivery_date) %>% 
   slice(1) %>% 
   ungroup() %>% 
   distinct(patid, delivery_date)
 

#bind together
 total_gdm_single_deliveries <- bind_rows(gdm_obs_outside_del_single_round2, gdm_pregnancy_search_single, in_pregnancy_gdm_single)
 
```
Now lastly because, in the second round we ensure that we were looking for episodes that had been missed because of a likely erroneous 1st jan delivery date picked up in the first round, we need to prioritise the TRUE delivery date
```{r}
 total_gdm_single_deliveries_slimmed <- total_gdm_single_deliveries %>%
     # ---- Rule: identify 01-01 placeholders ----
 mutate(
   year   = year(delivery_date),
   is_jan1 = month(delivery_date) == 1 & day(delivery_date) == 1
 ) %>%
   group_by(patid, year) %>%
   filter(!(is_jan1 & n() > 1)) %>%   # drop Jan 1 when another date exists
   ungroup() %>%
      # ---- Rule: create 9-month clusters ----
 arrange(patid, delivery_date) %>%
   group_by(patid) %>%
   mutate(
     cluster = cumsum(c(0, diff(delivery_date) > months(9)))
   ) %>%
      # ---- Rule: pick 1 date in each cluster ----
 group_by(patid, cluster) %>%
   mutate(
     has_id       = any(!is.na(delivery_id)),
     cluster_size = n(),
     keep = case_when(
       # If any delivery_id exists → keep ONLY the row with ID because that was created above
       has_id ~ !is.na(delivery_id),
              # No ID and multiple dates → keep the latest
       !has_id & cluster_size > 1 ~ delivery_date == max(delivery_date),
              # No ID and single date → keep it
       !has_id & cluster_size == 1 ~ TRUE
     )
   ) %>%
   filter(keep) %>%
   ungroup() %>% 
   mutate(gdm_del_date = delivery_date) %>% 
   distinct(patid, delivery_date, delivery_id, gdm_del_date) 
 
```

Some other potentially relevant delivery data

```{r}

 ###all deliveries of women loged with gdm OTHER than the deliveries following a pregnancy complicated by GDM
  other_deliveries <- unique_deliveries_single %>% 
   filter(patid %in% total_gdm_single_deliveries_slimmed$patid) %>% 
   anti_join(total_gdm_single_deliveries_slimmed, by = c("patid", "delivery_date")) %>% #remove the gdm delivery dates
   distinct(patid, delivery_date, delivery_id)
 
  #all the deliveries for gdm patients, included those the pregnancies that were not coded with GDM
 all_gdm_patients_deliveries_single <- bind_rows(other_deliveries, total_gdm_single_deliveries_slimmed)
 saveRDS(all_gdm_patients_deliveries_single, "all_gdm_patients_deliveries_single.rds")
 
 #for all unique deliveries for single patid (whether or not they had GDM), which were NOT coded with GDM
 all_deliveries_minus_gdm_episodes <- unique_deliveries_single %>% 
   anti_join(total_gdm_single_deliveries_slimmed, by = c("patid", "delivery_date")) %>% 
   distinct(patid, delivery_date, delivery_id)
 
 #for all unique deliveries for single patids, get ALL their deliveries plus GDM episodes suggestive of pregnancies (albeit without a delivery date)
 all_deliveries_single <-bind_rows(all_deliveries_minus_gdm_episodes, total_gdm_single_deliveries_slimmed) %>% 
   left_join(first_obs_gly_drug_HES %>% select(patid, yob), by = "patid") %>% 
   filter(year(delivery_date)>yob & (year(gdm_del_date)>yob)|is.na(gdm_del_date))
 
```

Now do the same process as above with those with multiple patids
```{r}
 unique_deliveries_multiple <- multiple(unique_deliveries) %>% 
   rename(delivery_date = obsdate)
 
 unique_deliveries_multiple <- unique_deliveries_multiple %>%
   mutate(
     start = delivery_date %m-% months(9),  # 9 months before delivery
     end   = delivery_date %m+% months(3)   # 3 months after delivery
   )
 
 in_pregnancy_multiple <- all_pregnancy_obs_update %>%
   inner_join(unique_deliveries_multiple %>% select(individual_id, start, end),
              by = "individual_id") %>% ##expect many to many
   filter(obsdate >= start & obsdate <= end)
 
 gdm_obs_outside_del_multiple <-multiple(all_pregnancy_obs_update) %>%
   anti_join(in_pregnancy_multiple %>% select(individual_id, obsdate, start, end), by = c("individual_id", "obsdate")) %>% #expect many to many
   filter(any(gdm == 1)) %>%
   filter(obsdate < earliest_preexisting %m-% years(1)|is.na(earliest_preexisting)) %>% 
   filter(!grepl("history", Term, ignore.case = TRUE)) %>% 
   filter(!grepl("h/o", Term, ignore.case = TRUE)) %>% 
   distinct()
 
 #exmining these it is clear that there are some gdm episodes that do actually ahve a delivery date but, for example, it was misdate as 01-01 or there is a historical delivery observation during the pregnancy that is conflicting
 # Having removed the delivery date obtained above, if we cycle through to find more delivery dates that should remove some ambiguity
 
 unique_deliveries_gdm_multiple <- gdm_obs_outside_del_multiple %>%
   filter(!is.na(delivery)) %>% 
   arrange(individual_id, obsdate) %>%
   group_by(individual_id) %>%
   mutate(
     obsdate = as.Date(obsdate),
     gap_from_prev = as.numeric(obsdate - lag(obsdate)),
     # Create cluster IDs for episodes within 2 months (<= 60 days)
     new_cluster = if_else(is.na(gap_from_prev) | gap_from_prev > 60, 1L, 0L),
     cluster_id = cumsum(new_cluster)
   ) %>%
   ungroup() %>% 
   # Collapse each cluster to the earliest observation date
   group_by(individual_id, cluster_id) %>%
   slice_min(order_by = obsdate, n = 1, with_ties = FALSE) %>%
   ungroup() %>%
   arrange(individual_id, obsdate) %>%
   group_by(individual_id) %>%
   mutate(
     # Apply the 9-month rule between clustered delivery dates
     gap_from_prev_delivery = as.numeric(obsdate - lag(obsdate)),
     new_delivery = if_else(is.na(gap_from_prev_delivery) | gap_from_prev_delivery >= 273, 1L, 0L),
     delivery_id = cumsum(new_delivery)
   ) %>%
   ungroup() %>% 
   group_by(individual_id, delivery_id) %>% 
   slice_min(order_by = obsdate, n = 1, with_ties = FALSE) %>% #then pick the earliest date within each delivery_id
   ungroup()
 
 unique_deliveries_gdm_multiple <- unique_deliveries_gdm_multiple %>%
   rename(delivery_date = obsdate) %>% 
   mutate(
     start = delivery_date %m-% months(9),  # 9 months before delivery
     end   = delivery_date %m+% months(3)   # 3 months after delivery
   )
 
 #find further pregnancies associated with a delivery as defined by the second round
 in_pregnancy_gdm_multiple <- gdm_obs_outside_del_multiple %>%
   inner_join(unique_deliveries_gdm_multiple %>% select(individual_id, start, end),#expect many to many
              by = "individual_id") %>%
   filter(obsdate >= start & obsdate <= end) %>% 
   distinct(individual_id, obsdate, start, end)
 
 #now find all gdm data outside those deliveries
 gdm_obs_outside_del_multiple_round2 <- gdm_obs_outside_del_multiple %>%
   anti_join(in_pregnancy_gdm_multiple %>% select(individual_id, obsdate, start, end), by = c("individual_id", "obsdate")) %>% 
   filter(!is.na(gdm)) %>% 
   group_by(individual_id) %>% 
   slice_min(obsdate) %>% 
   slice(1) %>% 
   ungroup() %>% 
   rename(delivery_date = obsdate) %>% 
   distinct(individual_id, delivery_date)
 
 in_pregnancy_gdm_multiple <- in_pregnancy_gdm_multiple %>% 
   rename(delivery_date=obsdate) %>% 
   distinct(individual_id, delivery_date)
 
 total_gdm_multiple_deliveries <- bind_rows(gdm_obs_outside_del_multiple_round2,  gdm_pregnancy_search_multiple, in_pregnancy_gdm_multiple)
 
 total_gdm_multiple_deliveries_slimmed <- total_gdm_multiple_deliveries %>%
   # ---- Rule: identify 01-01 placeholders ----
 mutate(
   year   = year(delivery_date),
   is_jan1 = month(delivery_date) == 1 & day(delivery_date) == 1
 ) %>%
   group_by(individual_id, year) %>%
   filter(!(is_jan1 & n() > 1)) %>%   # drop Jan 1 when another date exists
   ungroup() %>%
   # ---- Rule: create 9-month clusters ----
 arrange(individual_id, delivery_date) %>%
   group_by(individual_id) %>%
   mutate(
     cluster = cumsum(c(0, diff(delivery_date) > months(9)))
   ) %>%
   # ---- Rule: pick 1 date in each cluster ----
 group_by(individual_id, cluster) %>%
   mutate(
     has_id       = any(!is.na(delivery_id)),
     cluster_size = n(),
     keep = case_when(
       # If any delivery_id exists → keep ONLY the row with ID
       has_id ~ !is.na(delivery_id),
       # No ID and multiple dates → keep the latest
       !has_id & cluster_size > 1 ~ delivery_date == max(delivery_date),
       # No ID and single date → keep it
       !has_id & cluster_size == 1 ~ TRUE
     )
   ) %>%
   filter(keep) %>%
   ungroup() %>%
      mutate(gdm_del_date = delivery_date) %>% 
   distinct(individual_id, delivery_date, delivery_id, gdm_del_date)
  
 
 ##same for multiple
 other_deliveries_multiple <- unique_deliveries_multiple %>% 
   filter(individual_id %in% total_gdm_single_deliveries_slimmed$patid) %>% 
   anti_join(total_gdm_single_deliveries_slimmed, by = c("patid", "delivery_date")) %>% 
   distinct(patid, delivery_date, delivery_id)
 
 all_gdm_patients_deliveries_multiple <- bind_rows(other_deliveries, total_gdm_multiple_deliveries_slimmed)
 saveRDS(all_gdm_patients_deliveries_multiple, "all_gdm_patients_deliveries_multiple.rds")
 
 all_deliveries_minus_gdm_episodes_multiple <- unique_deliveries_multiple %>% 
   anti_join(total_gdm_multiple_deliveries_slimmed, by = c("individual_id", "delivery_date")) %>% 
   distinct(individual_id, delivery_date, delivery_id)
 
 all_deliveries_multiple <-bind_rows(all_deliveries_minus_gdm_episodes_multiple, total_gdm_multiple_deliveries_slimmed)
 
```

