---
title: "demographics"
author: "Gabrielle Goldet"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#some basic libraries
```{r}
library(dplyr)
library(lubridate)
library(ggplot2)
library(tidyr)
library(stringr)
options(scipen = 999)
```

#some basic functions
```{r}
#filtering according to whether an individual has or doesn't have more than 1 patid
single <- function(df){df %>% filter(is.na(individual_id))}
multiple <- function(df){df %>% filter(!is.na(individual_id))}

#propagating relevant data across all the patids of a given individual
fill_columns <- function(df, ...) {
  df %>% 
    left_join(all_pats, by = "individual_id") %>%
    group_by(individual_id) %>%
    fill(..., .direction = "downup") %>%
    ungroup()
}

#finding the earliest date across events for all an individual's patids, taking into account some or none of them may have had it (avoids error messages because of NAs)
safe_min <- function(x) if (all(is.na(x))) NA_Date_ else min(x, na.rm=TRUE)

#filter down to earliest obsdate per patid or individual_id
obsdate_single_fun <- function(df){df %>% group_by(patid) %>% slice_min(obsdate) %>% slice(1) %>% ungroup()}
obsdate_multiple_fun <- function(df){df %>% group_by(individual_id) %>% slice_min(obsdate) %>% slice(1) %>% ungroup()}
```



## R Creating age at diagnosis 
gdm_full contains data on the GDM group. Age at diagnosis was calculated by deducting the year of birth from the year at diagnosis and is stored in the age_at_diagnosis column.
non_gdm_full contains datain on non-GDM group 1 (i.e. the total non-GDM group)
filtered is the dataframe produced by 6_final_cohort_formation.rmd containing the data on the final total cohort, with GDM status and date of diabetes diagnosis
```{r}
IQR(gdm_full$age_at_diagnosis)

hist(gdm_full$age_at_diagnosis)
median(gdm_full$age_at_diagnosis)
non_gdm_full <- filtered %>% 
  filter(is.na(gdm_status))

IQR(non_gdm_full$age_at_diagnosis)
hist(non_gdm_full$age_at_diagnosis)
median(non_gdm_full$age_at_diagnosis) 

#prepare df 'age' for hypothesis testing
age <-gdm_full %>% 
  select(age_at_diagnosis) %>% 
  mutate(group = 1) %>% #label the GDM group
  bind_rows(., non_gdm_full %>% select(age_at_diagnosis)) %>% 
  mutate(group = if_else(is.na(group), 2, group),#label the non-GDM group
         group = as.factor(group))
#hypothesis test
wilcox.test(age_at_diagnosis~group, data= age)

#barplot for interest
ggplot(age, aes(x = age_at_diagnosis, fill = group)) +
  geom_histogram(aes(y = after_stat(density)),
                 position = "identity",
                 alpha = 0.4,
                 binwidth = 2) +
  theme_classic() +
  labs(x = "Age",
       y = "Density",
       fill = "Group")
```
the above can be repeat by comparing with other non-GDM groups (e.g. the age-matched group 2 or the age- and pregnancy matched group 3).


## ethnicity
First the ethnicity coding in secondary care was used where available.  This is stored in hes_patient as gen_ethnicity.
all_pats is a df containing the patids of those individuals with multiple patids along with their unique individual_id

```{r}
#add on the individual_id
hes_patient<- hes_patient %>% 
  left_join(all_pats, by = "patid")

# need to treat patients with single and multiple IDs differencly
gdm_single_full <- gdm_full %>% 
  filter(!is.na(patid)) %>%  
  left_join(hes_patient %>% select(patid, gen_ethnicity), by = "patid") %>% 
  filter(!is.na(gen_ethnicity)) %>% 
  distinct(patid, gen_ethnicity)

gdm_multiple_full <- multiple(gdm_full) %>% 
  distinct(individual_id) %>% 
  left_join(hes_patient %>% select(individual_id, gen_ethnicity), by = "individual_id") %>% 
  filter(!is.na(gen_ethnicity)) %>% 
  distinct(individual_id, gen_ethnicity)

#check to see if there are any individuals who have multiple ethnicities recorded in HES
test <- gdm_multiple_full %>% 
  group_by(individual_id) %>% 
  filter(n_distinct(gen_ethnicity)>1)
#in our case no discordant ethnicities

#bind all the gdm patients together
gdm_ethnicity <- bind_rows(gdm_single_full, gdm_multiple_full)

#determine the gdm patients with sincle patids lacking an ethnicity
ethnicity_to_find_gdm_single <- single(gdm_full) %>%
  filter(!patid %in% gdm_single_full$patid)

#determine the gdm patients with multiple patids lacking an ethnicity
ethnicity_to_find_gdm_multiple <- multiple(gdm_full) %>%
  filter(!individual_id %in% gdm_multiple_full$individual_id) %>% 
  select(-patid) %>% 
  fill_columns()

#repeat the process above for the non-gdm group
non_gdm_single_full <- single(non_gdm_full) %>% 
  left_join(hes_patient %>% select(patid, gen_ethnicity), by = "patid") %>% 
  filter(!is.na(gen_ethnicity)) %>% 
  distinct(patid, gen_ethnicity)

non_gdm_multiple_full <- multiple(non_gdm_full) %>% 
  distinct(individual_id) %>% 
  left_join(hes_patient %>% select(individual_id, gen_ethnicity), by = "individual_id") %>% 
    filter(!is.na(gen_ethnicity)) %>% 
  distinct(individual_id, gen_ethnicity)

test <- non_gdm_multiple_full %>% 
  group_by(individual_id) %>% 
  filter(!is.na(gen_ethnicity)) %>% 
  filter(n_distinct(gen_ethnicity)>1)
#no idscordant ethnicities

non_gdm_ethnicity <- bind_rows(non_gdm_single_full %>% select(patid, gen_ethnicity), non_gdm_multiple_full %>%  select(individual_id, gen_ethnicity))

#the below creates a df with the counts and percentages of each ethnicity
percentages_non_gdm_full <- as.data.frame(table(non_gdm_ethnicity$gen_ethnicity)) %>% 
  mutate(percentage = Freq*100/sum(Freq))

#identify those without an ethnicity in HES
ethnicity_to_find_single <- single(non_gdm_full) %>%
  filter(!patid %in% non_gdm_single_full$patid)

ethnicity_to_find_multiple <- multiple(non_gdm_full) %>% 
  filter(!individual_id %in% non_gdm_multiple_full$individual_id) %>% 
  select(-patid) %>% 
  fill_columns() %>% #this brings in the original patids (necessary for the enxt step)
  distinct(patid, individual_id)

#build total df of subjects whose ethnicity is missing
ethnicity_to_find <- bind_rows(ethnicity_to_find_multiple, ethnicity_to_find_single, ethnicity_to_find_gdm_multiple, ethnicity_to_find_gdm_single) %>% 
  distinct(patid, individual_id)
```
Now go find missing ethnicities in cprd
```{r}

setwd("..") #set your working directory
ethnicity_code <- read.table("ethnicity_code.txt", header = TRUE, sep = "\t") #this is the code list available in same github repository
ethnicity_code$medcodeid <- as.character(ethnicity_code$medcodeid)

#function to extract all ethnicity related data (per code present on the ethnicity code list) on the subjects of interes
ethnicity_fun <- function(df){df %>% 
    inner_join(ethnicity_to_find, by = "patid") %>%  #just select the patients we are interested in
    left_join(ethnicity_code, by = "medcodeid") %>% #join ethnicity code
    filter(!is.na(term))} #only keep data related to ethnicity noting that the term column contains the name of the ethnicity

ethnicity_obs <- bind_rows(lapply(character_list, ethnicity_fun))


```

The cprd data is more complicated and any given patient may have more than one ethnic category (note all the ethnicities have been split into the 5 census categories).  The most commonly occurring ethnicity is selected in the process below

```{r}
#for those with single patids
ethnicity_single <- single(ethnicity_obs) %>% 
  distinct(patid, term, eth5) %>% 
    filter(eth5!=5) %>% 
  group_by(patid, eth5) %>%
  summarise(n = n(), .groups = "drop_last") %>%   # count occurrences
  slice_max(order_by = n, n = 1, with_ties = FALSE) %>%  # pick most frequent code per person
  ungroup() %>%
  select(patid, eth5)

#for those with multiple patids
ethnicity_multiple <- multiple(ethnicity_obs) %>% 
  distinct(individual_id,eth5) %>% 
  filter(eth5!=5) %>% 
  group_by(individual_id, eth5) %>%
  summarise(n = n(), .groups = "drop_last") %>%   # count occurrences
  slice_max(order_by = n, n = 1, with_ties = FALSE) %>%  # pick most frequent code per person
  ungroup() %>%
  select(individual_id, eth5)

##split into GDM/non-GDM groups
ethnicity_single_gdm <- ethnicity_single %>% 
  inner_join(filtered_single %>% select(patid, gdm_status), by = "patid") %>% 
  filter(gdm_status=="GDM")

ethnicity_single_non_gdm <- ethnicity_single %>% 
  inner_join(filtered_single %>% select(patid, gdm_status), by = "patid") %>%
  filter(is.na(gdm_status))

ethnicity_multiple_gdm <- ethnicity_multiple %>% 
  inner_join(filtered_multiple %>% select(individual_id, gdm_status), by = "individual_id") %>% 
  filter(gdm_status=="GDM")

ethnicity_multiple_non_gdm <- ethnicity_multiple %>% 
  inner_join(filtered %>% select(individual_id, gdm_status), by = "individual_id") %>% 
  filter(is.na(gdm_status))
```

now make totalethnicities for each group, and complete the eth5 variable from the CPRD data with the gen_ethnicty from the HES data
```{r}
gdm_ethnicity_full <- bind_rows(ethnicity_single_gdm, gdm_single_full, ethnicity_multiple_gdm, gdm_multiple_full) %>% 
  select(-gdm_status) %>% 
  mutate(eth5 = case_when(gen_ethnicity=="White"~0, #the gen_ethnicity comes from the HES data and needs to be added into the eth5 category
                       gen_ethnicity=="Chinese"|gen_ethnicity=="Indian"|gen_ethnicity=="Oth_Asian"|gen_ethnicity=="Pakistani"|gen_ethnicity=="Bangladesi"~1,
                        gen_ethnicity=="Bl_Carib"|gen_ethnicity=="Bl_Afric"|gen_ethnicity=="Bl_Other"~2,
                          gen_ethnicity=="Other"~3,
                          gen_ethnicity=="Mixed"~4,
                          TRUE~eth5 )) %>% 
  select(patid, individual_id, gen_ethnicity, eth5)

gdm_ethnicity_full <-gdm_ethnicity_full %>% 
   select(patid, individual_id, gen_ethnicity, eth5)%>% 
  mutate(group = "GDM")

table(gdm_ethnicity_full$eth5)

prop.table(table(gdm_ethnicity_full$eth5))
table(gdm_ethnicity_full$eth5, useNA = "ifany")

#do the same for the non-gdm group
non_gdm_ethnicity_full <- bind_rows(ethnicity_single_non_gdm, non_gdm_single_full, ethnicity_multiple_non_gdm, non_gdm_multiple_full) %>% 
  select(-gdm_status) %>% 
    mutate(eth5 = case_when(gen_ethnicity=="White"~0,
                       gen_ethnicity=="Chinese"|gen_ethnicity=="Indian"|gen_ethnicity=="Oth_Asian"|gen_ethnicity=="Pakistani"|gen_ethnicity=="Bangladesi"~1,
                          gen_ethnicity=="Bl_Carib"|gen_ethnicity=="Bl_Afric"|gen_ethnicity=="Bl_Other"~2,
                          gen_ethnicity=="Other"~3,
                          gen_ethnicity=="Mixed"~4,
                          TRUE~eth5)
         )


non_gdm_ethnicity_full <- bind_rows(non_gdm_ethnicity_full) %>% 
   select(patid, individual_id, gen_ethnicity, eth5) %>% 
  mutate(group = "NGDM")


prop.table(table(non_gdm_ethnicity_full$eth5))
table(non_gdm_ethnicity_full$eth5, useNA = "ifany")

prop.table(table(gdm_ethnicity_full$eth5))
table(gdm_ethnicity_full$eth5, useNA = "ifany")

#now X2 test for hypothesis testing

table_eth <- bind_rows(gdm_ethnicity_full, non_gdm_ethnicity_full) %>% 
  select(patid, individual_id, group, eth5)
 table_eth <- table(table_eth$group, table_eth$eth5)
 table_eth
chisq.test(table_eth)
```
Again this can be repeated with other non-GDM groups

#IMD
With CPRD data, one receives a .txt file with colums for pracid and e_2019_imd_10 (the practice-level IMD). This
```{r}
#read in your pracid - IMD file
setwd("...")
practice_imd <- read.delim("....txt", stringsAsFactors = TRUE)

#read in the patient file which links patids to pracids
setwd("...")
patient_file <- read.delim("...", stringsAsFactors = TRUE)

practice_imd_patient <- patient_file %>% 
  select(patid, pracid, regstartdate, regenddate) %>% #just keep relevant columns
  left_join(practice_imd, by = "pracid")

#prepare variables
practice_imd_patient$patid <- as.character(practice_imd_patient$patid)
practice_imd_patient$regenddate <- dmy(practice_imd_patient$regenddate)
practice_imd_patient$regstartdate <- dmy(practice_imd_patient$regstartdate)

#start with adding in the imd to the GDM group and just keeping patients with an IMD
gdm_imd_single <- single(gdm_full) %>% 
  left_join(practice_imd_patient, by = "patid") %>% 
  filter(!is.na(e2019_imd_10))

# for those with multiple patids the practid that was closest to the point of diagnosis of diabetes (stored in the new_ddd column) is chosen, then if there are two dates that are similarly distributed around the diabetes diagnosis date, select the registration that is current (i.e. regeenddate is NA) and if there are still multiple lines per patient chose the longest registration as stored in the time_in_surgery column
gdm_imd_multiple <- multiple(gdm_full) %>%
  select(-patid, -regstartdate, -regenddate) %>% 
  left_join(all_pats, by = "individual_id") %>% #brings patids back
  left_join(practice_imd_patient %>% select(patid, e2019_imd_10, regstartdate, regenddate), by = "patid") %>%  #this brings in the regstart date 
    select(patid, individual_id, e2019_imd_10, regstartdate, regenddate, new_ddd) %>% 
    filter(regstartdate<=new_ddd %m-% years(1)) %>% 
  filter(!is.na(e2019_imd_10)) %>% 
  mutate(time_in_surgery = if_else(!is.na(regenddate), as.numeric(regenddate-regstartdate), as.numeric(as.Date("2024-06-01")-regstartdate))) %>% 
  mutate(closest_date = abs(as.numeric(new_ddd-regstartdate)),
         priority = if_else(is.na(regenddate), 1, 2)) %>% 
  group_by(individual_id) %>% 
  slice_min(closest_date) %>% 
  slice_min(priority) %>% 
  slice_max(time_in_surgery) %>% 
  slice(1) %>% ##for a small number of individuals there are two identical registration periods at different pracids with diff IMD, pick one at random in the slice 1 above
  ungroup()

#now bind together all the gdm patients together and label the group as GDM
gdm_imd_full <- bind_rows(gdm_imd_single %>% select(patid, individual_id, e2019_imd_10), gdm_imd_multiple %>% select(patid, individual_id, e2019_imd_10)) %>% 
  mutate(group = "GDM")

#calculate percentages for each IMD group for the GDM group
percentages_gdm<- as.data.frame(table(gdm_imd_full$e2019_imd_10)) %>% 
  mutate(Percentage = Freq*100/sum(Freq),
         group = "GDM")%>% 
  rename(IMD_decile = Var1)
summary(gdm_imd_full)

non_gdm_imd_single <- non_gdm_full %>% 
  filter(!is.na(patid)) %>% 
  left_join(practice_imd_patient, by = "patid") %>% 
   filter(!is.na(e2019_imd_10)) 

non_gdm_imd_multiple <-multiple(non_gdm_full) %>%
  select(-patid, -regstartdate, -regenddate) %>% 
  left_join(all_pats, by = "individual_id") %>% 
  left_join(practice_imd_patient, by = "patid") %>% 
  select(patid, individual_id, e2019_imd_10, regstartdate, regenddate, new_ddd) %>% 
  filter(regstartdate<=new_ddd %m-% years(1)) %>% 
  filter(!is.na(e2019_imd_10)) %>% 
   mutate(closest_date = abs(as.numeric(new_ddd-regstartdate)),
          time_in_surgery = if_else(!is.na(regenddate), as.numeric(regenddate-regstartdate), as.numeric(as.Date("2024-06-01")-regstartdate)),
          priority = if_else(is.na(regenddate), 1, 2)) %>% 
  group_by(individual_id) %>% 
 slice_min(closest_date) %>% 
  slice_min(priority) %>% 
  slice_max(time_in_surgery) %>% 
  slice(1) %>% 
  ungroup()

non_gdm_imd <- bind_rows(non_gdm_imd_single %>% select(patid, individual_id, e2019_imd_10), non_gdm_imd_multiple %>% select(patid, individual_id, e2019_imd_10)) %>% 
  mutate(group = "Non-GDM")


#calculate percentages for each IMD group for the non-GDM group
percentages_ngdm<- as.data.frame(table(non_gdm_imd$e2019_imd_10)) %>% 
  mutate(Percentage = Freq*100/sum(Freq),
         group = "Non-GDM") %>% 
  rename(IMD_decile = Var1)
summary(non_gdm_imd)

#calc numbers of patients with missing IMD
nrow(gdm_full)-nrow(gdm_imd_full)
nrow(non_gdm_full) - nrow(non_gdm_imd)

```
This can be repeated for the other non-gdm groups


total pracids for whom we have an IMD
```{r}
patient_file$patid <- as.character(patient_file$patid)

pracid_filtered <- patient_file %>% 
  select(patid, pracid) %>% 
  left_join(all_pats)

single_pracid_filtered <- single(pracid_filtered) %>% 
  filter(patid %in% filtered_single$patid)

multiple_pracid_filtered <- multiple(pracid_filtered) %>% 
  filter(individual_id %in% filtered_multiple$individual_id)

pracid_filtered <- bind_rows(single_pracid_filtered, multiple_pracid_filtered)

#total number of pracids in cohort
n_distinct(pracid_filtered$pracid)


practices_with_imd <- practice_imd_patient %>% 
  left_join(all_pats)

single_practice_imd <-practices_with_imd %>% 
  filter(patid %in% filtered_single$patid)
multiple_practice_imd <- multiple(practices_with_imd) %>% 
  filter(individual_id %in% filtered_multiple$individual_id)

filtered_with_imd <- bind_rows(single_practice_imd, multiple_practice_imd) %>% 
  filter(!is.na(e2019_imd_10))

n_distinct(filtered_with_imd$pracid)

summary(df_log_reg)
```


#now for BMI
note that "character_list" is a list of dataframes containing raw data extracted by cprd. A practical solution to the large aamount of data received was to remove columns that I was not going to use and combine the many files into (in my case 17) different data frames (rather than 1 df that would have been too large to load). I kept the columns for patid, medcodeid, obsdate and value. Having loaded each of these 17 dataframes, I created a list of these dataframes (termed character_list, as I had converted the patids from numerical to character) and with that I was able to use functions and lapply to execute various operations on all the data.
```{r}
#load all relevant code lists available in the github repository
setwd("...") #your working directory
bmi_code <- readRDS("bmi_code.rds")
height <- read.csv("height.csv", header = TRUE)
weight <- read.csv("weight.csv", header = TRUE)

#now get BMI data from the character_list of dataframes containing raw cprd data
add_bmi_fun <- function(df){df %>% left_join(bmi_code, by = "medcodeid") %>% filter(!is.na(term))}
bmi_obs <- bind_rows(lapply(character_list, add_bmi_fun))

#format date
bmi_obs$obsdate <-dmy(bmi_obs$obsdate)

#add the individual_id for those with multiple patids
bmi_obs <- bmi_obs %>% 
    left_join(all_pats) 

#split into single and multiple and find the bmi closest to the diabetes diagnosis date (new_ddd)  
bmi_obs_single <- bmi_obs %>% 
  inner_join(filtered_single %>% select(patid, new_ddd), by = "patid") %>% 
  mutate(from_ddd = abs(as.numeric(obsdate-new_ddd))) %>% 
  filter(value>14) %>% 
  filter(value<60) %>% 
   group_by(patid) %>% 
  arrange(patid, from_ddd) %>% 
  slice(1) %>% 
  ungroup()


bmi_obs_multiple <- multiple(bmi_obs) %>%
    inner_join(filtered_multiple %>% select(individual_id, new_ddd), by = "individual_id") %>% 
  mutate(from_ddd = abs(as.numeric(obsdate-new_ddd))) %>% 
  filter(value>14) %>% 
  filter(value<60) %>% 
 group_by(patid) %>% 
  arrange(patid, from_ddd) %>% 
  slice(1) %>% 
  ungroup()

##get quantitative height and weight observations
height_weight_code <- bind_rows(height, weight)
height_weight_code$medcodeid <- as.character(height_weight_code$medcodeid)

add_height_weight_fun <- function(df){df %>% 
    inner_join(height_weight_code, by = "medcodeid") %>% 
    filter(!is.na(obsdate))}

height_weight_obs <- bind_rows(lapply(character_list, add_height_weight_fun)) %>% 
  left_join(all_pats)#add individual_id

height_weight_obs <- height_weight_obs %>% 
  filter(!is.na(value) & !is.na(obsdate)) %>% #only keep rows containing a value and an observation date
  select(patid, obsdate, medcodeid, value, term, height, comment, weight, individual_id) 

#split into single and multiple patid dfs
height_weight_obs_single <- single(height_weight_obs) %>% 
 left_join(filtered_single %>% select(patid, new_ddd, yob), by = "patid") %>% 
  filter(year(obsdate)>yob+17) %>% #only keep observations in adulthood
  filter(value>0)

#find height per patient
closest_height_single <- height_weight_obs_single %>% 
  filter(!is.na(height)) %>% 
  filter(value>100 & value <220) %>% #remove errouneous heights
      mutate(from_ddd = abs(as.numeric(obsdate-new_ddd))) %>% #find gap between that recording tand diabetes diagnosis
  group_by(patid) %>% 
   slice_min(from_ddd, n = 1, with_ties = FALSE) %>% #select height closet to diabetes diagnosis
  ungroup()

#calculate closest weight to diagnosis of DM
closest_weight_single <- height_weight_obs_single %>% 
    mutate(from_ddd = abs(as.numeric(obsdate-new_ddd))) %>% 
  filter(!is.na(weight)) %>% 
  filter(value>35 & value<250) %>% #data clean
  select(-weight) %>% 
  group_by(patid) %>% 
  slice_min(from_ddd, n = 1, with_ties = FALSE) %>% #find the weight closest to the diabetes diagnosis 
  ungroup() %>% 
  rename(weight_date = obsdate,
         weight = value)

#now do those with multiple patids
height_weight_obs_multiple <- multiple(height_weight_obs) %>% 
 left_join(filtered %>% select(individual_id, new_ddd, yob), by = "individual_id") %>% 
  filter(year(obsdate)>yob+17) %>% 
  filter(value>0)

#find height per patient
closest_height_multiple <- height_weight_obs_multiple %>% 
  filter(!is.na(height)) %>% 
  filter(value>100 & value <220) %>% 
      mutate(from_ddd = abs(as.numeric(obsdate-new_ddd))) %>% 
  group_by(individual_id) %>% 
 slice_min(from_ddd, n = 1, with_ties = FALSE) %>% 
  ungroup()

#calculate closest weight to diagnosis of DM
closest_weight_multiple <- height_weight_obs_multiple %>% 
    mutate(from_ddd = abs(as.numeric(obsdate-new_ddd))) %>% 
  filter(!is.na(weight)) %>% 
   filter(value>35 & value<250) %>% 
  select(-weight) %>% 
  group_by(individual_id) %>% 
  slice_min(from_ddd, n = 1, with_ties = FALSE) %>% 
  ungroup() %>% 
  rename(weight_date = obsdate,
         weight = value)

#now add in HES obesity coding
# all HES apc data is stored in all_HES_apc

all_HES_apc$patid <- as.character(all_HES_apc$patid)

#load BMI code list for HES
setwd("...") # yourdirectory
bmi_icd <- read.delim("bmi_icd.txt", header = TRUE, sep = "\t") #codelist on github

HES_bmi <- all_HES_apc %>% 
  left_join(all_pats, by = "patid") %>% #add indivudyak_id
  inner_join(bmi_icd, by = "ICD") %>% #keep only rows related to BMI
  select(patid, individual_id, admidate, ICD, term) %>% 
  mutate(admidate = as.Date(admidate, format = "%Y/%m/%d"), #format dates
         obese = 1) %>% #flag all this data as relating to obesity
  rename(obsdate = admidate) #rename the date to obsdate so that it can be lined up wiht the cprd data

#filter by those present in the cohort 
HES_bmi_single <- single(HES_bmi) %>% 
  filter(patid %in% filtered$patid)

HES_bmi_multiple <- multiple(HES_bmi) %>% 
  filter(individual_id %in% filtered$individual_id)

##now put it all together
calc_bmi_single <- closest_height_single %>% 
  select(-weight,  -individual_id) %>% #remove unnecessary columns
  inner_join(closest_weight_single %>% select(patid, weight_date, weight), by = "patid") %>% 
  mutate(height_m = value/100,  #convert to metres
    bmi_value = weight/(height_m^2)) %>%  #calculate bmi
  filter(bmi_value < 60) %>% # get rid of bmis that are erroneous
  mutate(obsdate = coalesce(weight_date, obsdate)) %>% #merge the dates
    bind_rows(.,bmi_obs_single%>% select(patid, obsdate, term, new_ddd, bmi, low_bmi, normal_bmi, high_bmi, obese, value)) %>% 
  bind_rows(., HES_bmi_single %>% select(patid, obsdate, term, obese)) %>% 
  mutate(bmi_value = if_else(!is.na(bmi), value, bmi_value)) %>% #bring in the value from bmi_obs
   mutate(from_ddd = abs(as.numeric(obsdate-new_ddd))) %>% 
  group_by(patid) %>% 
  slice_min(from_ddd) %>% #get the closest date to the diabetes diagnosis date
  slice(1) %>% 
  ungroup() %>% 
  left_join(filtered_single %>% select(patid, gdm_status), by = "patid") #bring in the gdm status

#same for those with multiple patids
calc_bmi_multiple <- closest_height_multiple %>% 
  select(-weight, -weight_4cat, -patid) %>% 
  inner_join(closest_weight_multiple %>% select(individual_id, weight_date, weight), by = "individual_id") %>% 
  mutate(height_m = value/100,
    bmi_value = weight/(height_m^2)) %>% 
  filter(bmi_value < 60) %>% 
  mutate(obsdate = coalesce(weight_date, obsdate)) %>% 
  bind_rows(.,bmi_obs_multiple%>% select(individual_id, obsdate, term, new_ddd, bmi, low_bmi, normal_bmi, high_bmi, obese, value)) %>% 
   bind_rows(., HES_bmi_multiple %>% select(individual_id, obsdate, term, obese)) %>% 
  mutate(bmi_value = if_else(!is.na(bmi), value, bmi_value)) %>% 
   mutate(from_ddd = abs(as.numeric(obsdate-new_ddd))) %>% 
  group_by(individual_id) %>% 
  slice_min(from_ddd) %>% 
  slice(1) %>% 
  ungroup() %>% 
  left_join(filtered_multiple %>% select(individual_id, gdm_status), by = "individual_id")
  
##group these into bmi categories
calc_bmi_single <- calc_bmi_single%>% 
  mutate(bmi_value = coalesce(bmi_value, value), #some values for bmi that had not been picked up earlier
    bmi_cat = case_when(bmi_value<18.5|!is.na(low_bmi)~"Underweight",
                             bmi_value>=18.5&bmi_value<25|!is.na(normal_bmi)~"Normal weight",
                             bmi_value>= 25&bmi_value<30| !is.na(high_bmi)~"Overweight",
                             bmi_value>=30 & bmi_value<40| !is.na(obese) ~"Obese",
                        bmi_value>40~"Severely obese"
                        )) 

calc_bmi_multiple <- calc_bmi_multiple%>% 
  mutate(bmi_value = coalesce(bmi_value, value),
    bmi_cat = case_when(bmi_value<18.5|!is.na(low_bmi)~"Underweight",
                             bmi_value>=18.5&bmi_value<25|!is.na(normal_bmi)~"Normal weight",
                             bmi_value>= 25&bmi_value<30| !is.na(high_bmi)~"Overweight",
                             bmi_value>=30| !is.na(obese) ~"Obese",
                        bmi_value>40~"Severely obese")) 

table(calc_bmi_single$bmi_cat)

barplot(calc_bmi_single$bmi_cat)


#now for calculation of raw quantitative BMI
  BMI_raw <- bind_rows(calc_bmi_multiple, calc_bmi_single) %>% 
    select(patid, individual_id, gdm_status, bmi_value) %>% 
    mutate(gdm_status = if_else(is.na(gdm_status), 0, 1)) %>% 
    filter(!is.na(bmi_value)) %>% 
    group_by(gdm_status) %>% 
    summarise(mean_bmi = mean(bmi_value),
              median_bmi = median(bmi_value),
              IQR_bmi = IQR(bmi_value))

  hist(BMI_raw$bmi_value) #examine distribution of BMIs
  
    BMI_raw <- bind_rows(calc_bmi_multiple, calc_bmi_single) %>% 
    select(patid, individual_id, gdm_status, bmi_value) %>% 
    mutate(gdm_status = if_else(is.na(gdm_status), 0, 1)) %>% 
    filter(!is.na(bmi_value))
    
##now get GDM and non gem single and multiple

gdm_bmi_single <- calc_bmi_single %>% 
  filter(!is.na(gdm_status)) %>% 
  select(patid, bmi_cat, bmi_value)

non_gdm_bmi_single <- calc_bmi_single %>% 
  filter(is.na(gdm_status)) %>% 
  select(patid, bmi_cat, bmi_value)

gdm_bmi_multiple <-  calc_bmi_multiple %>% 
  filter(!is.na(gdm_status)) %>% 
  select(individual_id, bmi_cat, bmi_value)

non_gdm_bmi_multiple <- calc_bmi_multiple %>% 
  filter(is.na(gdm_status)) %>% 
  select(individual_id, bmi_cat, bmi_value)

#bind all the gdm data
gdm_bmi <- bind_rows(gdm_bmi_single, gdm_bmi_multiple) %>% 
  mutate(group = "GDM")

percentages <- as.data.frame(table(gdm_bmi$bmi_cat)) %>% 
  mutate(percentages = Freq*100/sum(Freq))

non_gdm_bmi <- bind_rows(non_gdm_bmi_single, non_gdm_bmi_multiple) %>% 
  mutate(group = "Non-GDM")

percentages <- as.data.frame(table(non_gdm_bmi$bmi_cat)) %>% 
  mutate(percentage = Freq*100/sum(Freq))

#calc the number of subjects with missing BMI
nrow(non_gdm_full)-nrow(non_gdm_bmi)

nrow(gdm_full)-nrow(gdm_bmi)

##now for distributions
hist(gdm_bmi$bmi_value) # slight right skew so use wilcoxon

hist(non_gdm_bmi$bmi_value)
```

now for p and some summary statistics
```{r}
table_bmi <- bind_rows(non_gdm_bmi, gdm_bmi) %>% 
  select(patid, individual_id, group, bmi_cat) 
 table_bmi <- table(table_bmi$group, table_bmi$bmi_cat)
 
chisq.test(table_bmi)

#medians
median(gdm_bmi$bmi_value, na.rm = TRUE)
IQR(gdm_bmi$bmi_value, na.rm = TRUE)

median(non_gdm_bmi$bmi_value, na.rm = TRUE)
IQR(non_gdm_bmi$bmi_value, na.rm = TRUE)


bmi <- bind_rows(gdm_bmi, non_gdm_bmi)
wilcox.test(bmi_value~gdm_status, data = bmi)

```

#smoking
Again the character list of dfs containing all observations is necessary here.
The character drugs list is simiar to the character list but instead of observations, itcontains drug issue data
```{r}
setwd("....") #your dir
smoking_drugs <- read.delim("smoking_meds.txt", stringsAsFactors = TRUE)
smoking_code <-readRDS("smoking_code.rds")

smoking_drugs <- smoking_drugs %>% 
    rename(prodcodeid = ProdCodeId,
         productname = ProductName) %>% 
    mutate(productname = if_else(productname=="", TermfromEMIS, productname)) %>% 
  select(prodcodeid, productname)

#create a function to extract smoking data from character list 
smoking_fun <- function(df){df %>% left_join(smoking_code, by = "medcodeid") %>% filter(!is.na(term))}

#extract smoking observations
smoking_obs <-bind_rows(lapply(character_list, smoking_fun)) %>% 
  left_join(all_pats, by = "patid") #add the individual id


#this one code was identified as coming with a value that looked like the number of cigarettes a day and 0 was assumed to mean non-smoker"
smoking_obs <- smoking_obs %>% 
    mutate(ever_smoker = case_when(medcodeid=="108938018" & value == 0~ 0, #if smokes no cigarettes, assume long-term non-smoker
                                   medcodeid=="108938018" & value != 0~ 1, #if non-zero, assume smoked at some time
                                   TRUE~ever_smoker),#or keep columns as they are
         smoking_status = case_when(medcodeid=="108938018" & value == 0~ 0, #if smokes no cigarettes, assume non-smoker
                                   medcodeid=="108938018" & value != 0~ 2, #if non-zero, assume current smoker
                                   TRUE~smoking_status))
 
smoking_obs$obsdate <- dmy(smoking_obs$obsdate)

#just patients with single-Patids
filtered_single <- single(filtered)

#add individual_id
smoking_obs <- smoking_obs %>% 
  left_join(all_pats, by = "patid")

#now filter to relevant patients
smoking_single <- smoking_obs %>% 
  inner_join(filtered_single%>%select(patid, gdm_status, new_ddd, yob), by = "patid") %>%  
  filter(year(obsdate)>yob+8) %>% #get reasonable date
  filter(obsdate<=new_ddd) %>% 
  group_by(patid) %>% 
    mutate(ever_updated = if_else(any(smoking_status!=0), 1, 0)) %>% #for each patient identify any eveidence of smoking at any point
  slice_max(obsdate) %>% #take the closest date before new_ddd
  slice(1) %>% 
  mutate(ever_updated2 = pmax(smoking_status, ever_updated)) %>% #take the max of the smoking status and whether they have ever smoked.  This ensures that any current smoking is logged as an ex-smoker
ungroup()

#note smoking_multiple below

##now add in drug cessation data
smoking_drugs_fun <- function(df){df %>% inner_join(smoking_drugs, by = "prodcodeid") %>% filter(!is.na(issuedate))}

smoking_drugs_obs <- bind_rows(lapply(character_drugs_list, smoking_drugs_fun)) %>% 
  left_join(all_pats)

smoking_drugs_obs_single <- single(smoking_drugs_obs) %>% 
  inner_join(filtered_single %>% select(patid, gdm_status, new_ddd, yob), by = "patid") %>%  
  filter(year(obsdate)>yob+8) %>% 
  filter(obsdate<=new_ddd) %>% 
  group_by(patid) %>% 
    mutate(ever_updated = if_else(any(smoking_status!=0), 1, 0)) %>% #for each patient identify any evidence of smoking at any point
  slice_max(obsdate) %>% #take the closest date before new_ddd
  slice(1) %>% 
  mutate(ever_updated2 = pmax(smoking_status, ever_updated))

filtered_multiple <- multiple(filtered)

smoking_drugs_obs_multiple <- multiple(smoking_drugs_obs) %>% 
    inner_join(filtered_multiple %>% select(individual_id, gdm_status, new_ddd, yob), by = "individual_id")

smoking_total_single<- bind_rows(smoking_single, smoking_drugs_obs_single) %>% 
  filter(year(obsdate)>yob+8) %>% 
  filter(obsdate<=new_ddd) %>% 
  group_by(patid) %>% 
    mutate(ever_updated = if_else(any(smoking_status!=0), 1, 0)) %>% #for each patient identify any evidence of smoking at any point
  slice_max(obsdate) %>% #take the closest date before new_ddd
  slice(1) %>% 
  mutate(ever_updated2 = pmax(smoking_status, ever_updated))

##now split into GDM and not
gdm_smoking_single <- smoking_total_single %>% 
  filter(!is.na(gdm_status))

gdm_single <- single(gdm_full)
non_gdm_single <- single(non_gdm_full)

missing_gdm_smoking_single <-gdm_single %>% 
  filter(!patid %in% gdm_smoking_single$patid) #39

non_gdm_smoking_single <- smoking_total_single %>% 
  filter(is.na(gdm_status))

missing_non_gdm_smoking_single <- non_gdm_single %>% 
  filter(!patid %in% non_gdm_smoking_single$patid) #449

##now for the multiple patid - this coding could be cleaned up and all the filtering down ot the bound dfs


smoking_obs_multiple <-smoking_obs %>%  
  inner_join(filtered_multiple%>%select(individual_id, gdm_status, new_ddd, yob), by = "individual_id") %>%  
  filter(year(obsdate)>yob+8) %>% 
  filter(obsdate<=new_ddd) %>% 
  group_by(individual_id) %>% 
    mutate(ever_updated = if_else(any(smoking_status!=0), 1, 0)) %>% #for each patient identify any evidence of smoking at any point
  slice_max(obsdate) %>% #take the closest date before new_ddd
  slice(1) %>% 
  mutate(ever_updated2 = pmax(smoking_status, ever_updated)) %>% #take the max of the smoking status and whether they have ever smoked.  This ensures that any current smoking is logged as an ex-smoker
ungroup()
  

smoking_total_multiple <- bind_rows(smoking_obs_multiple, smoking_drugs_obs_multiple) %>% 
  filter(year(obsdate)>yob+8) %>% 
  filter(obsdate<=new_ddd) %>% 
  group_by(individual_id) %>% 
    mutate(ever_updated = if_else(any(smoking_status!=0), 1, 0)) %>% #for each patient identify any evidence of smoking at any point
  slice_max(obsdate) %>% #take the closest date before new_ddd
  slice(1) %>% 
  mutate(ever_updated2 = pmax(smoking_status, ever_updated)) %>% #take the max of the smoking status and whether they have ever smoked.  This ensures that any previous current smoking is logged as an ex-smoker
  ungroup()


gdm_multiple <- multiple(gdm_full)
  
gdm_smoking_multiple <- smoking_total_multiple %>% 
  filter(!is.na(gdm_status))

missing_gdm_smoking_multiple <-gdm_multiple %>% 
  filter(!individual_id %in% gdm_smoking_multiple$individual_id) #10

gdm_smoking_single <- smoking_total_single %>% 
  filter(!is.na(gdm_status))

 
non_gdm_smoking_multiple <- smoking_total_multiple %>% 
filter(is.na(gdm_status))

non_gdm_multiple <- multiple(filtered) %>% 
  filter(is.na(gdm_status))

non_gdm_smoking_single <- smoking_total_single %>% 
  filter(is.na(gdm_status))

missing_non_gdm_smoking_multiple <- non_gdm_multiple %>% 
  filter(!individual_id %in% non_gdm_smoking_multiple$individual_id) 

gdm_smoking <- bind_rows(gdm_smoking_single, gdm_smoking_multiple) %>% 
  mutate(group = "GDM")
percentages <- as.data.frame(table(gdm_smoking$ever_updated2)) %>% 
  mutate(percentage = Freq*100/sum(Freq))

non_gdm_smoking <- bind_rows(non_gdm_smoking_single, non_gdm_smoking_multiple,) %>% 
  mutate(group = "Non-GDM")


percentages <- as.data.frame(table(non_gdm_smoking$ever_updated2)) %>% 
  mutate(percentage = Freq*100/sum(Freq))

#calc number with missing data
nrow(gdm_full)-nrow(gdm_smoking)
nrow(non_gdm_full)-nrow(non_gdm_smoking)
```

now for stats

```{r}

table_smoking <- bind_rows(non_gdm_smoking, gdm_smoking) %>% 
  select(patid, individual_id, group, ever_updated2)
 table_smoking <- table(table_smoking$group, table_smoking$ever_updated2)
 
 table_smoking
chisq.test(table_smoking)
```

#Follow-up time
```{r}
f_up_time <- filtered %>% 
  mutate(f_up = if_else(!is.na(regenddate)|!is.na(death_date), pmin(regenddate, death_date, na.rm = TRUE) - new_ddd, as.Date("2024-12-01") - new_ddd)) %>% 
  select(patid, individual_id, gdm_status,new_ddd, regenddate, death_date, f_up) %>% 
  mutate(f_up = as.numeric(stringr::str_extract(f_up, "\\d+")), # converts difftime object to plain number
         group = if_else(is.na(gdm_status), "NGDM", "GDM"))
  

gdm_f_up <- f_up_time %>% 
  filter(!is.na(gdm_status))
median(gdm_f_up$f_up)
hist(gdm_f_up$f_up)
IQR(gdm_f_up$f_up)

ngdm_f_up <- f_up_time %>% 
  filter(is.na(gdm_status))

median(ngdm_f_up$f_up)
hist(ngdm_f_up$f_up)
IQR(ngdm_f_up$f_up)

wilcox.test(f_up~group, data= f_up_time)
```

#calculate the date that what drives ddd
```{r}
filtered <- filtered %>% 
  mutate(driver = case_when(new_ddd==issuedate~"drug",
                            new_ddd==admidate~"HES",
                            new_ddd==ddd~"diag",
                            new_ddd==gly_date~"gly",
                            new_ddd==gly_date & new_ddd==ddd | new_ddd==gly_date & new_ddd==gly_date & new_ddd==issuedate |new_ddd==gly_date & new_ddd==admidate|new_ddd==issuedate & new_ddd==ddd| new_ddd==admidate & new_ddd==ddd|new_ddd==issuedate & new_ddd==admidate~"multiple"))
table(filtered$driver)
```

#get average HbA1C
need diagnostic_gly_obs, which is a large df containing all the observations related to codes on the gly_code list available on github
```{r}
average_hba1c_single <- filtered_single %>% 
  select(patid, new_ddd, gdm_status) %>% 
  left_join(diagnostic_gly_obs, by = "patid") %>% 
  filter(hba1c==1) %>% #get all observations flagged as relating to hba1c   
  select(patid, new_ddd, obsdate, value) %>%
  filter(value<200 & (obsdate>=new_ddd)) %>% #only keep values that are reasonable i.e. <200 and after diabetes diagnosis
  mutate(hba1c_mmol_mol = if_else(value>=48, value, (value - 2.15) * 10.929)) %>% #ensure all in same unit
  group_by(patid) %>% 
  summarise(average_hba1c = mean(hba1c_mmol_mol)) %>% 
    ungroup() %>% 
  left_join(filtered_single %>% select(patid, gdm_status), by = "patid")

average_hba1c_multiple <- filtered_multiple %>% 
  select(individual_id, new_ddd, gdm_status) %>% 
  left_join(diagnostic_gly_obs, by = "individual_id") %>% 
  filter(hba1c==1) %>%   
  select(individual_id, new_ddd, obsdate, value) %>%
  filter(value<200 & (obsdate>=new_ddd)) %>% 
  mutate(hba1c_mmol_mol = if_else(value>=48, value, (value - 2.15) * 10.929)) %>% 
  group_by(individual_id) %>% 
  summarise(average_hba1c = mean(hba1c_mmol_mol)) %>% 
    ungroup() %>% 
    left_join(filtered_multiple %>% select(individual_id, gdm_status), by = "individual_id")

total_average_hba1c <- bind_rows(average_hba1c_multiple, average_hba1c_single) %>% 
  mutate(gdm_status = if_else(is.na(gdm_status), "non GDM", gdm_status))

#check distribution type (in our case right skey)
hist(average_hba1c_single$average_hba1c)

table_hba1c <- total_average_hba1c %>% group_by(gdm_status) %>% 
  summarise(n = n(),
            median_hba1c = median(average_hba1c),
            iqr = IQR(average_hba1c),)
 
wilcox.test(average_hba1c ~ gdm_status, data = total_average_hba1c)
```